Lossy compression of DNA sequencing quality data

### 2016-05-05 bmc_manuscript/manuscript.tex

The economic costs of file size for binning, in general terms, include
no fixed costs, and marginal costs that are the product of the number
of base-call quality values times the cost of the encoding.

@@

The economic costs of file size for model-based compression, in
general terms, include no fixed costs, and marginal costs that are the
product of the number of reads times the cost of representing the
model parameters.

@@

The economic costs of file size for profile-based compression, in
general terms, include fixed costs associated with representing the
profiles, which is the product of the number of profiles times the
cost of encoding them. These fixed costs are amortized over the entire
set of reads to which they apply, and thus on a read basis vary as a
reciprocal function of the number of reads. Additionally there are
marginal costs that are the product of the number of reads encoded
times the cost of the encoding. Profiling like modeling, but not
binning, has the advantageous property that marginal costs are the
reciprocal of the read length. Thus the costs of file size for
profile-based compression decrease proportionally with increases in
either or both the number of reads and read length.

@@

For comparison to other developed methods for lossy compression of
quality values we analyzed the reference data with
\textsc{QualComp}~\cite{Ochoa:2013rt}, Read Quality Sparsifier
(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}, and Quality Values Zip
(\textsc{qvz)}~\cite{Malysa01102015}.

@@

The economic costs of file size for compression with \textsc{QualComp}
or \textsc{qvz}, in our interpretation, appear to be the fixed costs
for storing the cluster representatives, and the marginal costs that
are the product of the number of reads times the cost of representing
the model parameters.

\textsc{rqs} first constructs a dictionary of commonly occurring
$k$-mers from either the sequencing data set they seek to compress or
a larger population-sized sequencing data set. During compression,
$k$-mers within reads that have a small Hamming distance from a
dictionary $k$-mer are identified. The explicit assumption is that
positions differing from the $k$-mers in the dictionary more likely
correspond to SNPs or sequencing errors. These positions that differ
from the nearest $k$-mer in the dictionary are marked. Lastly, the
quality values of the remaining positions are smoothed (discarded? -
see paper).

The economic costs of file size for compression with \textsc{rqs}, in
our interpretation, appear to be the fixed costs of the dictionary of
commonly occurring $k$-mers, and the marginal costs that are the
product of the number of positions of $k$-mers within the chosen
Hamming distance from dictionary $k$-mers times the cost of
representing these positions. As with compression based on profiles,
amortization applies to fixed costs of compression with
\textsc{rqs}. Here the fixed costs of the dictionary of commonly
occurring $k$-mers can be amortized over reads for which the
dictionary may be appropriately used (e.g., reads from the same
species).

>>>>

The economic costs of file size for binning, in general terms, include
no fixed costs, and marginal costs that are the product of the number
of base-call quality values multiplied by the cost of the encoding.

@@

The economic costs of file size for model-based compression, in
general terms, include no fixed costs, and marginal costs that are the
product of the number of reads multiplied by the cost of representing
the model parameters.

@@

The economic costs of file size for profile-based compression, in
general terms, include fixed costs associated with representing the
profiles, which is the product of the number of profiles multiplied by
the cost of encoding them. These fixed costs are amortized over the
entire set of reads to which they apply, and thus on a read basis vary
as a reciprocal function of the number of reads. Additionally there
are marginal costs that are the product of the number of reads encoded
multiplied by the cost of the encoding. Profiling like modeling, but
not binning, has the advantageous property that marginal costs are the
reciprocal of the read length. Thus the costs of file size for
profile-based compression decrease proportionally with increases in
either or both the number of reads and read length.

@@

For comparison to other developed methods for lossy compression of
quality values we analyzed the reference data with
\textsc{QualComp}~\cite{Ochoa:2013rt}, and Quality Values Zip
(\textsc{qvz)}~\cite{Malysa01102015}, and Read Quality Sparsifier
(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}

@@

The economic costs of file size for compression with \textsc{QualComp}
or \textsc{qvz}, in our interpretation, appear to be the fixed costs
for storing the cluster representatives, and the marginal costs that
are the product of the number of reads multiplied by the cost of
representing the model parameters.

Compression with \textsc{rqs} first begins with constructing a
dictionary of commonly occurring $k$-mers from either the sequencing
data set to be compressed or a larger population-sized sequencing
data set. During the second phase, sparsification, $k$-mers within
reads that have a small Hamming distance from a dictionary $k$-mer are
identified. Base-calls within these $k$-mers are deemed high
confidence and the associated quality values discarded, as are quality
values above a chosen threshold for base-calls in other $k$-mers. The
explicit assumption is that positions differing from the $k$-mers in
the dictionary more likely correspond to SNPs or sequencing errors. At
a subsequent step the discarded quality values are replaced with the
threshold value, in a manner somewhat similar to the binning strategy
described previously.

The economic costs of file size for compression with \textsc{rqs}, in
our interpretation, appear to be the fixed costs of the dictionary of
commonly occurring $k$-mers, and the marginal costs that are the
product of the number of base-calls represented as high-quality or
other value multiplied by the cost of representing these values. As
with compression based on profiles, amortization applies to fixed
costs of compression with \textsc{rqs}. Here the fixed costs of the
dictionary of commonly occurring $k$-mers can be amortized over reads
for which the dictionary may be appropriately used (e.g., reads from
the same species).
